---
title: "Statistical Inf_p1"
author: "Aroogz"
date: "November 9, 2016"
output: html_document
---

```{r echo=FALSE}
options(warn = -1)
```
## SYNOPSIS: the analysis will investigate the exponential distribution in R and compare it with the Central Limit Theorem


## Preparing the Workspace
```{r preparing workspace, message=FALSE}
rm(list = ls())
if ("ggplot2" %in% row.names(installed.packages()) == FALSE){install.packages("ggplot2")}
if ("grid" %in% row.names(installed.packages()) == FALSE){install.packages("grid")}
if ("gridExtra" %in% row.names(installed.packages()) == FALSE){install.packages("gridExtra")}
library(ggplot2); library(grid); library(gridExtra)
```
## Investigating the Exponential Distribution

### get the cumulative means and variances
```{r }
mns <- NULL; vars <- NULL; lambda <- 0.2
# simulating 1000 means and variances
for (i in 1:1000){
  mns <- c(mns, mean(rexp(40, lambda)))
  vars <- c(vars, var(rexp(40, lambda)))
}
# getting the mean of the means
cummeans <- cumsum(mns)/(1:1000); cumvar <- cumsum(vars)/(1:1000)
```
### the theoretical values
```{r theoretical values}
mean.theor <- 1/lambda
var.theor <- (1/lambda)^2
print(paste("theoretical mean: ", mean.theor))
print(paste("theoretical variance: ", var.theor))
```
### the distribution of means converging to a normal
```{r comparing with normal}
set.seed(100)
#sample normal distribution
p1 <- ggplot(mapping = aes(rnorm(1000, mean = mean.theor)))+
  geom_histogram(binwidth = 0.2, col="red", fill= "blue", alpha=0.2)+
  labs(x = "", title= "Sample Normal Distribution")
#distribution of 1000 random exps; not gaussian
p2 <- ggplot(data.frame(val = rexp(1000, 0.2)), aes(val))+
  geom_histogram(binwidth = 0.8, col="red", fill= "blue", alpha=0.2)+
  labs(x="", title= "Dist. of 1000 random Samples")
# distribution of means
mns.data <- data.frame(mns)
p3 <- ggplot(mns.data, aes(mns))+ 
  geom_histogram(binwidth = 0.2, col="red", fill= "blue", alpha=0.2)+
  geom_vline(xintercept = c(mean.theor, mean(mns)), col= c("blue", "red"),
                            size = 0.5)+
  labs(x="", title= "Dist. of 1000 random Sample means")
# arranging plots
set.seed(100)
lay <- rbind(c(1,2),c(3,4))
grid.arrange(p1, p1, p2, p3, layout_matrix= lay,
             top = "checking obedience of the central limit theorem of the mean")
```
it is observable that the distribution of the means of the exponential distribution (bottom right) converges to a Normal distibution as compared the distribution of a 1000 exponentials on the bottom left. Thus showing the Central Limit Theorem at work.

### showing the convergence of the mean distribution
```{r showing convergence}
ggplot(mapping = aes(1:1000, cummeans))+
  geom_line()+
  geom_hline(yintercept = mean.theor, col= "red")+
  labs(x = "number of outcomes", y = "mean of the outcome",
       title= "convergence of the mean of exponential dist. to the theoretical value")
```
we see in the plot above as the mean distribution converges to the theoretical mean, `theo.mean` (marked by the horizontal line) as the outcomes increase

### the distribution of variances converging to normal
```{r comparing variance dist, with normal}
#sample normal distribution
p5 <- ggplot(mapping = aes(rnorm(1000, mean = var.theor)))+
  geom_histogram(binwidth = 0.2, col="red", fill= "blue", alpha=0.2)+
  labs(x = "", title= "Sample Normal Distribution")

# distribution of variances
p6 <- ggplot(mapping= aes(vars))+
  geom_histogram(binwidth = 0.2, col="red", fill= "blue", alpha=0.2)+
  geom_vline(xintercept = c(var.theor, mean(vars)), col= c("blue", "red"),
             size = 0.5)+
  labs(title= "Dist. of 1000 random Sample means", x= "")
set.seed(100)
lay2 <- rbind(c(1, 2),
              c(3, 4))
grid.arrange(p5,p5, p2, p6, layout_matrix= lay2,
             top= "checking obedience of the central limit theorem")
```
it is observable that the distribution of the variances of the exponential distribution (bottom right) converges to a Normal distibution as compared the distribution of a 1000 exponentials on the bottom left. Thus showing the Central Limit Theorem at work.

### showing the convergence of the mean variances
```{r convergence of the mean variances}
ggplot(mapping = aes(1:1000, cumvar))+
  geom_line()+
  geom_hline(yintercept = var.theor, col= "red")+
  labs(x = "number of outcomes", y = "variance of the outcomes",
       title= "convergence of the variance of exponential dist. to the theoretical value")
```
we see in the plot above as the mean distribution converges to the theoretical variance `theor.var` (marked by the horizontal line) as the outcomes increase.
