---
title: "Statistica Inf. P2"
author: "Aroogz"
date: "November 9, 2016"
output: html_document
---

## SYNOPSIS: will analyse the `ToothGrowth` dataset elucidating on the effect of `Supp` and `dose` variables on the `len`.

## Preparing the Workspace
```{r preparing workspace, message=FALSE}
rm(list = ls())
if ("ggplot2" %in% row.names(installed.packages()) == FALSE){install.packages("ggplot2")}
if ("grid" %in% row.names(installed.packages()) == FALSE){install.packages("grid")}
if ("gridExtra" %in% row.names(installed.packages()) == FALSE){install.packages("gridExtra")}
library(ggplot2); library(grid); library(gridExtra)
```

## 2. The analysis of the `ToothGrowth` dataset

The dataset consists of 60 rows and 3 columns. It looks intuitive that the observations are paired by the `supp` variable so we would consider this case.

### Visualising the data showing some plots
```{r dist according to supp}
ggplot(data = ToothGrowth)+
  geom_boxplot(mapping = aes(x=supp, y= len) ) +
  labs(title= "comparing the distributions for the two supplements disregarding the dose")
```
Here we observe the distribution of the `len` variable with respect to the `supp` variable

```{r}
# var explained by dose, grid by supp
ggplot(ToothGrowth)+
  geom_point(mapping = aes(x=factor(dose), y=len, col= factor(dose)))+
  facet_grid(supp~.)+
  labs(title= "variatons explained by the dosage for each of the supplement")
```
It is obvious from the above, the effect of the `dose` on the outcome of the `len` for both categories of `supp` variable

```{r vis. all factors}
#exploring variations due to all factors
ggplot(ToothGrowth)+
  geom_point(mapping = aes(x=factor(dose), y=len, col= supp))+
  labs(title= "variatons explained by the dose and supplement")
```
from the above plot, we observe an increase in the OJ `supp` for the first two dose (0.5 and 1). It becomes unclear the effect of the different supplements
### Assuming paired observation
```{r loading data}
data("ToothGrowth"); n <- nrow(ToothGrowth)/2

#divison alongn supp variable
g1 <- ToothGrowth$len[1:30]; g2 <- ToothGrowth$len[31:60]

# showing the confidence interval (Assuming paired observations)
mn <- mean(g2-g1); s <- sd(g2-g1)
mn + c(-1, 1)*qt(0.975, n-1)*s/sqrt(n)
```
we see the interval (95% confidence interval) here does not contain zero which implies that we could confidently reject the hypothesis of a zero mean difference. This implies that the variability due to the `supp` variable is indeed significant and could not have happened by chance.
### Assuming unpaired observation
```{r unpaired}
t.test(x= g2, y = g1, paired = FALSE)$conf
```
the unpaired test however shows an interval that contains zero and so we do not have enough evidence to reject the hypothesis of a 0 mean difference

### implementing the permutation method
```{r permutation}
#implement the permutation test
cal.diff <- function(value, group){
  # function to get the mean different for a list
  # differentiated by groups (another column)
  divs <- levels(group)
  meandiff <- mean(value[group == divs[1] ]) - mean(value[group == divs[2]])
  meandiff
}
observation <- cal.diff(ToothGrowth$len, ToothGrowth$supp)
permutations <- NULL
for (i in 1:1000){
  permutations <- c(permutations, cal.diff(ToothGrowth$len, 
                                           sample(ToothGrowth$supp)))
}
print(paste("the observed mean difference: ", observation))
```


```{r}
### check fraction of permutation greater than the observation
mean(permutations > observation)
```
we see a low percentage of the permutations greater than the observation
```{r permutations interval}
quantile(permutations, c(0.025,0.975))
```
we see it is highly unlikely to have gotten this value if it were left to randomness.
```{r obs. p-value}
quantile(permutations, 0.95)
```
the 95% quantile is also less than the observation indication our observation is very unlikely due to chance. hence there is a difference in the `len` is associated with the `supp`. Particularly, the mean difference is of the `len` is not zero in favour of the `OJ` `supp` category.
```{r clear, echo=FALSE}
rm(list = ls())
```

&copy; Aroge